---
title: "Analysis of Gazi bay eDNA data"
output: html_notebook
---


Load these packages:

library(vegan)
library(ggplot2)
library(dplyr)
library(cowplot)
library(ggplot2)
library(corrplot)
```{r packages, include=FALSE}
library(vegan)
library(ggplot2)
library(dplyr)
library(cowplot)
library(ggplot2)
library(corrplot)
```

Upload the Project_data into the R environment.
## Subset data
I labelled species data as "df" and metadata as "st_df"
```{r subset}
df<- data[,9:71] 
st_df <- data[,1:8] # subset columns
```

## Community diversity
### Alpha diversity
Shannon's H'
```{r H}
H <- diversity(df)
```

Observed Richness
```{r richness}
richness <- specnumber(df)
```

Pielou's Evenness
```{r Pielou}
evenness <- H/log(richness)
```

Create alpha diversity dataframe including environmental data
```{r alpha}
alpha <- cbind(shannon = H, richness = richness, pielou = evenness, st_df)
list(alpha)
```

Plot Alpha Diversity
```{r alpha plot, echo=FALSE}
plot.shan <- ggplot(alpha, aes(x =Site , y = shannon, colour = Site, shape = Site)) +
  geom_point(size = 3) +
  scale_colour_viridis_d(option = "A", begin = 0.2, end = 0.8) +
  ylab("Shannon's H'") + 
  xlab("") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.4))

plot.rich <-ggplot(alpha, aes(x = Site, y = richness, colour = Site, shape = Site)) +
  geom_point(size = 3) +
  scale_colour_viridis_d(option = "A", begin = 0.2, end = 0.8) +
  ylab("Species Richness") +
  xlab("Site") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.4))

plot.even <- ggplot(alpha, aes(x = Site, y = evenness, colour = Site, shape = Site)) +
  geom_point(size = 3) +
  scale_colour_viridis_d(option = "A", begin = 0.2, end = 0.8) +
  ylab("Pielou's Evenness") +
  xlab("") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.4))

#legend <- get_legend(plot.even)

plot_grid(plot.shan + theme(legend.position = "none"), plot.rich + theme(legend.position = "none"), plot.even + theme(legend.position = "none"),ncol = 3)

```

Save pdf
```{r}
pdf("~/Desktop/eDNA_221124/R_eDNA/Gazi-eDNA/figures/alpha-eDNA_sites.pdf",
    width = 12,
    height = 8,
    pointsize = 8)
plot_grid(plot.shan + theme(legend.position = "none"), plot.rich + theme(legend.position = "none"), plot.even + theme(legend.position = "none"),ncol = 3)
```

## BETA DIVERSITY

convert Replicates to factor to allow shapes number expansion from the minimun 6
```{r}
st_df$Replicate <- factor(st_df$Replicate)
```

## Calculate pairwise dissimilarity
Bray-Curtis Dissimilarity
```{r Bray dissimilarity}
data.mdf <- as.matrix.data.frame(df)
rownames(data.mdf) <- st_df$Replicate

data.bray <- vegdist(data.mdf, Replicate = "bray")
data.bray
```

Jaccard Dissimilarity
```{r Jaccard dissimilarity}
data.jac <- vegdist(data.mdf, Replicate = "jaccard", binary = T)
data.jac
```

# ORDINATION
## Principal Coordinates Analysis (PCoA)

calculate principal coordinates analysis (Bray-Curtis)
```{r PCoA}
pcoa.data.bray <- cmdscale(data.bray, k = 5, eig = T)
```
extract axis positions for each Replicate from cmdscale object and create a dataframe for plotting
```{r axis position}
pcoa.data.bray.plotting <- as.data.frame(pcoa.data.bray$points)
colnames(pcoa.data.bray.plotting) <- c("axis_1", "axis_2")
pcoa.data.bray.plotting$Replicate <- rownames(pcoa.data.bray.plotting)
```
calculate the proportion of variance in the data which is explained by the first two PCoA axes
```{r variance proportions}
pcoa.data.bray$eig[1]/(sum(pcoa.data.bray$eig))

pcoa.data.bray$eig[2]/(sum(pcoa.data.bray$eig))
```
Create a PCoA plot
```{r}
pcoa.data.bray.plot <- ggplot(pcoa.data.bray.plotting, aes(x = axis_1, y = axis_2, colour = Replicate)) +
  geom_point(size = 3) +
  scale_colour_viridis_d(option = "A", begin = 0.2, end = 0.8) +
  theme_bw() + 
  xlab("PCoA 1 (34.6%)") +
  ylab("PCoA 2 (16.9%)") +
  annotate(geom = 'text', label = 'Bray-Curtis', x = Inf, y = -Inf, hjust = 1.15, vjust = -1)
```

and repeat the above process with Jaccard's dissimilarity matrix

```{r}
pcoa.data.jac <- cmdscale(data.jac, k = 5, eig = T)

pcoa.data.jac.plotting <- as.data.frame(pcoa.data.jac$points)
colnames(pcoa.data.jac.plotting) <- c("axis_1", "axis_2")
pcoa.data.jac.plotting$Replicate <- rownames(pcoa.data.jac.plotting)

pcoa.data.jac$eig[1]/(sum(pcoa.data.jac$eig))

pcoa.data.jac$eig[2]/(sum(pcoa.data.jac$eig))
```


```{r}
pcoa.data.jac.plot <- ggplot(pcoa.data.jac.plotting, aes(x = axis_1, y = axis_2, colour= Replicate)) +
  scale_shape_manual(values=1:nlevels(pcoa.data.jac.plotting$Replicate)) +
  geom_point(size = 3) +
  scale_colour_viridis_d(option = "A", begin = 0.2, end = 0.8) +
  theme_bw() + 
  xlab("PCoA 1 (34%)") +
  ylab("PCoA 2 (20.8%)") +
  annotate(geom = 'text', label = 'Jaccard', x = Inf, y = -Inf, hjust = 1.215, vjust = -1)
```

Create Bray-Curtis PCoA and Jaccard PCoA plots side by side
```{r PCoA plot, echo=FALSE}
legend <- get_legend(pcoa.data.jac.plot) # extract plot legend
plot_grid(pcoa.data.bray.plot + theme(legend.position = 'none'), pcoa.data.jac.plot + theme(legend.position = 'none'), legend, ncol = 3, rel_widths = c(1,1,0.5))

```
save PDF
```{r}
pdf("~/Desktop/eDNA_221124/R_eDNA/Gazi-eDNA/figures/beta-eDNA_sites.pdf",
    width = 12,
    height = 8,
    pointsize = 8)
plot_grid(pcoa.data.bray.plot + theme(legend.position = 'none'), pcoa.data.jac.plot + theme(legend.position = 'none'), legend, ncol = 3, rel_widths = c(1,1,0.5))
```


Create a matrix
I converted my abundance data frame into a matrix and labelled it "m_data"
```{r matrix}
m_data <- as.matrix(df)
```

Screen for the best NMDS dimension
```{r dimension screening, message=FALSE, warning=FALSE}
# First step is to calculate a distance matrix. See PCOA for more information about the distance measures
# Here we use bray-curtis distance, which is recommended for abundance data
dist <- vegdist(m_data,  method = "bray")

# In this part, we define a function NMDS.scree() that automatically 
# performs a NMDS for 1-10 dimensions and plots the no. of dimensions vs the stress
NMDS.scree <- function(dist) { #where x is the name of the data frame variable
  plot(rep(1, 10), replicate(10, metaMDS(dist, autotransform = F, k = 1)$stress), xlim = c(1, 10),ylim = c(0, 0.30), xlab = "Number of Dimensions", ylab = "Stress", main = "NMDS stress plot")
  for (i in 1:10) {
    points(rep(i + 1,10),replicate(10, metaMDS(dist, autotransform = F, k = i + 1)$stress))
  }
}
# Use the function that we just defined to choose the optimal no. of dimensions
NMDS.scree(dist)
```


2-5 Dimensions show the best stress level which is below 0.2 and above 0. For this study I picked 4D.

Generate NMDS and view it;
```{r nmds}
set.seed(102)
nmds <- metaMDS(m_data, k = 4, trymax = 100, distance = "bray")
nmds
```
Calling your nmds object in R, will give you some information about your analysis. An important number to note is the stress, which is roughly the “goodness of fit” of your NMDS ordination. For a good representation of your data, the stress value should ideally be less than 0.2. If the stress value is 0, it might mean you have an outlier sample that is very different than all your other samples. Depending on your question, you may want to remove this sample to observe any other underlying patterns in your data.Stress values >0.2 are generally poor and potentially uninterpretable, whereas values <0.1 are good and <0.05 are excellent, leaving little danger of misinterpretation. Stress values between 0.1 and 0.2 are useable but some of the distances will be misleading. The stress value should be reported somewhere in your figure or figure caption. My stress value for this analysis is 0.02372954.

Create a Stressplot
```{r stressplot}
stressplot(nmds)
```

Create a simple NMDS plot in base R;
```{r basic NMDS}
plot(nmds)
```
The plot does not give any meaningful information. The circles represent samples while the red stars represent species.
From here we will be using **ggplot2** for nice figures;

First, I obtained the coordinates for NMDS1 and NMDS2 axes and put them in a new data frame (data.scores)
```{r}
#extract NMDS scores (x and y coordinates)
data.scores <- as.data.frame(scores(nmds, display = c("site"), tidy = T))
```

then I added columns with environmental data (Site, Replicate and Habitat) to my new data frame(data.scores).
```{r}
#add columns to data frame 
data.scores$Site = st_df$Site
data.scores$Replicate = st_df$Replicate
data.scores$Habitat = st_df$Habitat
 
head(data.scores)
```

Create an NMDS plot using ggplot2
```{r NMDS-ggplot2}
library(ggplot2)

siteplotnmds<- ggplot(data.scores, aes(x = NMDS1, y = NMDS2)) + 
    geom_point(size = 3, aes( shape = Site, colour = Habitat))+ 
    theme(axis.text.y = element_text(colour = "grey30", size = 12), 
    axis.text.x = element_text(colour = "black", face = "bold", size = 12), 
    legend.text = element_text(size = 12, face ="bold", colour ="grey30"), 
    legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), 
    axis.title.x = element_text(face = "bold", size = 14, colour = "black"), 
    legend.title = element_text(size = 14, colour = "black", face = "bold"), 
    panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2),
    legend.key=element_blank()) + 
    labs(x = "NMDS1", colour = "Habitat", y = "NMDS2", shape = "Site")  + 
    scale_colour_viridis_d(alpha = 1, begin = 0, end = 1, direction = 1, option = "D", aesthetics = "colour") 
siteplotnmds
```

From the plot I observed a distinction in my sample replicates and site with some replicates clustering at the center which is an indication that they contain some similar species.

```{r}
pdf("~/Desktop/eDNA_221124/R_eDNA/Gazi-eDNA/figures/NMDS-eDNA_sites.pdf",
    width = 12,
    height = 8,
    pointsize = 8)
siteplotnmds
```

First we need to extract the relevant information from nmds and envfit results.I extracted Sample coordinates in NMDS ordination space.
```{r}
#data.scores1 <- as.data.frame(scores(nmds, display = ("species"), tidy =T))
#data.scores1$Habitat <- st_df$Habitat
en = envfit(nmds, st_df, permutations = 999, na.rm = TRUE)
```

Extracting the required information from the envfit result is a bit more complicated. The envfit output contains information on the length of the segments for each variable. The segments are scaled to the r2 value, so that the environmental variables with a longer segment are more strongly correlated with the data than those with a shorter segment. You can extract this information with scores. Then these lengths are further scaled to fit the plot. This is done with a multiplier that is analysis specific, and can be accessed using the command ordiArrowMul(en). Below I multiply the scores by this multiplier to keep the coordinates in the correct proportion.
Because my data contained continuous and categorical environmental variables, I’m extracting the information from both separately using the “vectors” and “factors” options respectively.
```{r}
en_coord_cont <- as.data.frame(scores(en, "vectors")) * ordiArrowMul(en)
en_coord_cat <- as.data.frame(scores(en, "factors")) * ordiArrowMul(en)
```

To add species distribution to the plot,
I fit the species vector;
```{r spc vector}
data.spp.fit <- envfit(nmds, df, permutations = 999) # this fits species vectors
```

then transform species data into vectors and list p values
```{r spcs score}
spp.scrs <- as.data.frame(scores(data.spp.fit, display = "vectors")) #save species intrinsic values into dataframe
spp.scrs <- cbind(spp.scrs, Species = rownames(spp.scrs)) #add species names to dataframe
spp.scrs <- cbind(spp.scrs, pval = data.spp.fit$vectors$pvals) #add pvalues to dataframe so you can select species which are significant
spp.scrs<- cbind(spp.scrs, abrev = abbreviate(spp.scrs$Species, minlength = 6)) #abbreviate species names
sig.spp.scrs <- subset(spp.scrs, pval<=0.05) #subset data to show species significant at 0.05

spp.scrs
```
Now lets add some layer of species distribution and fit environmental parameters to the plot.
```{r final plot , echo=FALSE}
finalenv_spp_plot <- ggplot(data = data.scores, aes(x = NMDS1, y = NMDS2)) + 
     geom_point(data = data.scores, aes(colour = Habitat, shape = Site), size = 3, alpha = 0.5) + 
     scale_colour_viridis_d(alpha = 1,begin = 0,end = 1,direction = 1,option = "D",aesthetics = "colour")  + 
     geom_segment(aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2), 
       data = en_coord_cont, size =1, alpha = 0.5, colour = "grey30") +
     geom_text(data = en_coord_cont, aes(x = NMDS1, y = NMDS2), colour = "grey30", 
       fontface = "bold", label = row.names(en_coord_cont)) + 
     theme(axis.title = element_text(size = 10, face = "bold", colour = "grey30"), 
       panel.background = element_blank(), panel.border = element_rect(fill = NA, colour = "grey30"), 
       axis.ticks = element_blank(), axis.text = element_blank(), legend.key = element_blank(), 
       legend.title = element_text(size = 10, face = "bold", colour = "black"), 
       legend.text = element_text(size = 9, colour = "black")) + 
     labs(colour = "Habitat") +
 stat_ellipse(data= data.scores, aes(NMDS1, NMDS2, linetype = Habitat, colour = Habitat))+
   geom_point(data=spp.scrs, aes(x=NMDS1, y=NMDS2))

finalenv_spp_plot
```
```{r}
pdf("~/Desktop/eDNA_221124/R_eDNA/Gazi-eDNA/figures/finalNMDS-eDNA_sites_spp.pdf",
    width = 12,
    height = 8,
    pointsize = 8)
finalenv_spp_plot
```


## Statistical Tests

### ANOSIM
This tested whether there was a statistical difference between the fish community of each site and its replicate. The Null hypothesis: there is no difference between the fish communities in the samples.
ANOSIM is similar to ANOVA hypothesis test but uses a dissimilarity matrix as input instead of raw data. It is non-parametric which is perfect for my skewed eDNA fish abundance data. This complements the NMDS plot.

I used "anosim" function to test significant difference of my sample groups.
```{r sample anosim}
sig_test_Habitat<- anosim(df, st_df$Habitat, distance = "bray", permutations = 9999)
sig_test_Habitat
```

```{r habitat anosim}
sig_test_temp<- anosim(df, st_df$`Temperature(0C)`, distance = "bray", permutations = 9999)
sig_test_temp
```

```{r}
sig_test_ph<- anosim(df, st_df$`pH`, distance = "bray", permutations = 9999)
sig_test_ph
```

```{r}
sig_test_ppt<- anosim(df, st_df$`Salinity (ppt)`, distance = "bray", permutations = 9999)
sig_test_ppt
```

```{r}
sig_test_depth<- anosim(df, st_df$`Depth (m)`, distance = "bray", permutations = 9999)
sig_test_depth
```


### PERMANOVA

```{r PERMANOVA}
### Wisconsin double standardization then followed by Bray-Curtis dissimilarity matrix.
set.seed(100)
data.dist<- vegdist (wisconsin(df), "bray")

### Permutational multivariate analysis of variance-adonis. To test the significance of habitat difference in species distribution patterns.
adonis2(data.dist~st_df$Habitat)
adonis2(data.dist~st_df$`Temperature(0C)`)
adonis2(data.dist~st_df$`Depth (m)`)
adonis2(data.dist~st_df$`Salinity (ppt)`)
adonis2(data.dist~st_df$pH)
```
## Correlation Heatmap
Now create a correlation matrix with the community composition data using the command ‘cor’:
```{r cor}
cc <- cor(df, method = "spearman")
```

Pearson correlation: is the linear correlation between two variables.
Spearman correlation: is a non-parametric measure of rank correlation and assesses how well a relationship between two variables can be described using a monotonic function. A monotonic function is just a fancy way of describing a relationship where for each increasing x value, the y value also increases.
Plot;

##correlation with ggplot2
You need to manipulate the correlation matrix a bit to get it in the proper format for ggplot:
Change the matrix into a data frame format
Make the row names of your data frame into a column in your new data frame
Load in the package reshape2 for data manipulation
Finally, convert your "wide" format data frame into a "long" format data frame
```{r}
cc_df <- as.data.frame(cc)
cc_df$Species <- row.names(cc_df)
library(reshape2)
ccm <- melt(cc_df, id = "Species")
```

###Plot in ggplot
```{r}
corrplot <- ggplot(ccm, aes(x = variable, y = Species)) + 
    geom_tile(aes(fill = value), colour = "grey45") + 
    coord_equal() + 
    scale_fill_gradient2(low = "orange", mid = "white", high = "navy") + 
    theme(axis.text.y = element_text(face = "bold", colour = "grey25", size = 7), 
    legend.title = element_text(size = 10, face = "bold"),legend.position = "bottom", 
    axis.text.x = element_text(angle = 90, face = "bold.italic",colour = "grey25", vjust = 0.5, hjust = 0, lineheight = 5, size = 7), 
    panel.background = element_blank(), panel.border = element_rect(fill = NA, colour = NA), 
    axis.ticks = element_blank()) + 
    labs(x= "", y = "", fill = "Spearman's Correlation") + 
    scale_x_discrete(position = "top") +
    scale_y_discrete(limits = rev(levels(ccm$Species))) 
corrplot
```
Save PDF
```{r}
pdf("~/Desktop/eDNA_221124/R_eDNA/Gazi-eDNA/figures/corrplot.pdf",
    width = 12,
    height = 8,
    pointsize = 8)
corrplot
```






























